---
title: "Exercises for workshop: R for Linguistics"
date: "2026-06-01"
format:
  html:
    include-in-header: templates/head.html
    include-before-body: templates/before.html
    include-after-body: templates/after.html
    toc: true
    code-copy: hover
    link-external-newwindow: true

lang: en
---

```{=html}
<div class="lang-switcher">
  <a href="atelier_exercices_fr.html" title="Version française">🇫🇷</a>
  <span class="lang-label">FR</span>
</div>
```

```{r, echo=FALSE}
local({
  hook_source <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    x <- x[!grepl("^# *keep\\s*$", x)]
    hook_source(x, options)
  })
})
```

```{r, message=FALSE}
# keep
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## 2. Configuration

You must run the code below to check, and if necessary download and load, the
packages needed for the exercises.

```{r, message=FALSE}
# keep
source("funs.R")

verifier_packages(c("tidyverse", "glmnet", "tuneR", "phonTools", "umap", "dtw", "udpipe", "readtextgrid", "stringi"))
```

The code below configures the parameters needed for our work.

```{r, message=FALSE}
# keep
options(dplyr.summarise.inform = FALSE)
theme_set(theme_minimal())
Sys.setenv(LANG = "en")
```

## 3. Tabular Data

In the code block below, load the data
`donnees/peterson_barney_voyelle_eng.csv`, save the tibble in the object `phone`,
and inspect the data. Compare this table with the one in the notes.

```{r, message=FALSE}
phone <- read_csv2("donnees/peterson_barney_voyelle_eng.csv")
phone
```

## 4. Visualize

Create a visualization of the `phone` tibble showing F2 and F1 in the same form
as in the notes, but apply colour to indicate the group. If you have problems
with the symbols on your computer, use the `xsampa` column instead of IPA.

```{r}
phone |>
  ggplot() +
  geom_text(aes(x=f2, y=f1, label=api, colour=groupe)) +
  scale_x_reverse() +
  scale_y_reverse()
```

Create a boxplot of F1 values by IPA symbol using the `geom_boxplot` geometry.
Simply provide `x` and `y` aesthetics.

```{r}
phone |>
  ggplot() +
  geom_boxplot(aes(x=api, y=f1))
```

Ask questions if you do not understand the meaning of each part of a boxplot!

Create a boxplot visualization as above, but this time add a colour aesthetic
linked to the `groupe` column. Pay attention to the results.

```{r}
phone |>
  ggplot() +
  geom_boxplot(aes(x=api, y=f1, colour=groupe))
```

Copy the code above, then replace the `colour` aesthetic with `fill`.

```{r}
phone |>
  ggplot() +
  geom_boxplot(aes(x=api, y=f1, fill=groupe))
```

What is the difference between `colour` and `fill` in the grammar of graphics?

## 5. Manipulating a Table

Select the 5 rows with the highest F1 values.

```{r}
phone |>
  arrange(f1) |>
  slice_head(n=5)
```

Select the 5 rows with the lowest F1 values.

```{r}
phone |>
  arrange(desc(f1)) |>
  slice_head(n=5)
```

Select the rows that are *not* associated with men using the `!=` symbol instead
of `==`.

```{r}
phone |>
  filter(groupe != "homme")
```

Create new variables giving the difference between F1 and F0 and between F2 and
F1. These are sometimes useful for understanding the perceived qualities of
vowels.

```{r}
phone |>
  filter(groupe != "homme")
```

`quantile` is another function often applied inside `summarise`. You provide a
`probs` argument, which is a number between 0 and 1. The result is the value
below which `probs` percent of the data fall and above which `1-probs` percent
fall. For example, `quantile(f1, probs=0.5)` gives the same result as the
`median()` function. In the code below, create a table with a column `f1_q1`
giving the 0.25 quantile for each IPA symbol.

```{r}
phone |>
  group_by(api) |>
  summarise(
    f1_q1 = quantile(f1, probs=0.25)
  )
```

Continue by creating the columns `f1_q1`, `f1_q2`, `f1_q3`, `f2_q1`, `f2_q2`,
`f2_q3` with probs 0.25, 0.50, and 0.75. Save the results in a new object
`phone_quant`.

```{r}
phone_quant <- phone |>
  group_by(api) |>
  summarise(
    f1_q1 = quantile(f1, probs=0.25),
    f1_q2 = quantile(f1, probs=0.50),
    f1_q3 = quantile(f1, probs=0.75),
    f2_q1 = quantile(f2, probs=0.25),
    f2_q2 = quantile(f2, probs=0.50),
    f2_q3 = quantile(f2, probs=0.75)
  )
phone_quant
```

**Going further** There is a `geom_segment` geometry with four required
aesthetics: `x`, `xend`, `y`, and `yend`. This geometry creates line segments.
Create a visualization of F1 and F2 values using the quantiles.

```{r}
phone_quant |>
  ggplot() +
  geom_segment(aes(x=f2_q2, xend=f2_q2, y=f1_q1, yend=f1_q3)) +
  geom_segment(aes(x=f2_q1, xend=f2_q3, y=f1_q2, yend=f1_q2)) +
  geom_text(aes(x=f2_q2 - 50, y=f1_q2 - 25, label=api)) +
  scale_x_reverse() +
  scale_y_reverse()
```

## 6. Statistical Models

Load the data with the code below.

```{r, message=FALSE}
# keep
mots <- read_csv2("donnees/keylog-mots.csv.bz2")
mots
```

Create an object `mots_sub` from `mots` that contains only the words "the" and
"and".

```{r}
mots_sub <- mots |>
  filter(mot %in% c("the", "and"))
```

Perform a t-test to test the difference in duration between typing "the" and
typing "and".

```{r}
t.test(dur_mot ~ mot, data = mots_sub)
```

Perform a linear regression that predicts the duration of a word based on its
character length.

```{r}
summary(lm(dur_mot ~ char_mot, data = mots))
```

Interpret the results. Are they statistically significant?

## 7. Multiple Tables

Load the data with the code below.

```{r, message=FALSE}
meta <- read_csv2("donnees/keylog-meta.csv.bz2")
meta
```

Load another table with:

```{r}
langs <- read_csv2("donnees/langs.csv")
langs
```

Combine the `meta` and `langs` data using the `left_join` function.

```{r}
meta |>
  left_join(langs, by = "lang")
```

As described in the notes, combine `mots` and `meta`. Then combine with `langs`
and calculate the median of `d1`. Sort the rows by the medians.

```{r}
mots |>
  left_join(meta, by = "id") |>
  left_join(langs, by = "lang") |>
  group_by(groupe) |>
  summarise(mu = median(d1)) |>
  arrange(desc(mu))
```

## 8. Sliding Windows

Load the data with the code below.

```{r, message=FALSE}
touches <- read_csv2("donnees/keylog-touches.csv.bz2", na="NA")
touches
```

Use the `lead` function to create a new column `space_apres` indicating whether
the next key is a space. Then, use `filter` to select rows that are not spaces
themselves. Calculate the median keystroke duration based on whether a space
follows.

```{r}
touches |>
  mutate(space_apres = lead(code) == "Space") |>
  filter(code != "Space") |>
  filter(!is.na(space_apres)) |>
  group_by(space_apres) |>
  summarise(
    dur_moyenne = median(dur_apres)
  )
```

## 9. Character Strings

Load the data with the code below.

```{r, message=FALSE}
noms <- read_csv2("donnees/lexique_noms.csv")
noms
```

Use the `stri_extract` function to create a new column `ter` containing the last
character of each word. Calculate the proportion of feminine nouns for each
final letter, filter to letters with at least 10 words, and sort results in
descending order by proportion of feminine nouns.

```{r}
noms |>
  mutate(ter = stri_extract(lemme, regex="\\w\\Z")) |>
  group_by(ter) |>
  summarise(
    prop_f = mean(genre == "f"),
    n = n()
  ) |>
  filter(n > 10) |>
  arrange(desc(prop_f))
```

## 10. TextGrid

Load a Rhapsodie file with the code below.

```{r}
# keep
tg <- read_textgrid(
  "donnees/rhapsodie/tg/Rhap-D0001-Pro.TextGrid",
  encoding="UTF-8"
)
tg
```

Create an object `locuteur` containing only the "locuteur" tier.

```{r}
locuteur <- tg |>
  filter(tier_name == "locuteur")
locuteur
```

Create a summary counting the number of rows for each speaker value.

```{r}
locuteur |>
  group_by(text) |>
  summarise(n = n())
```

Using the `stri_count` function, create a new column `locuteur_nombre` in
locuteur that counts the number of "$" signs. Save the results using the arrow
symbol.

```{r}
locuteur <- locuteur |>
  mutate(locuteur_nombre = stri_count(text, fixed = "$"))
locuteur
```

Calculate the total duration for each speaker count.

```{r}
locuteur |>
  group_by(locuteur_nombre) |>
  summarise(
    duree = sum(xmax - xmin)
  )
```

## 11. Programming

The code below creates an object listing all files containing dialogues from
Rhapsodie.

```{r}
# keep
dir_nom <- "donnees/rhapsodie/tg/"
d <- dir(dir_nom, pattern="TextGrid$")
d <- d[stri_sub(d, 1, 6) == "Rhap-D"]
```

By modifying the code below, save the speaker information from each file,
including the number of speakers at each data point.

```{r}
df <- list("vector", length(d))
for (j in seq_along(d)) {
  tg <- read_textgrid(file.path(dir_nom, d[j]), encoding="UTF-8")
  tg$id <- j

  # ↓↓↓ cette partie ci-dessous fonctionne de traiter une seule
  # ↓↓↓ fiche ; vous pouvez la changer
  locuteur <- tg |>
    filter(tier_name == "locuteur") |>
    mutate(locuteur_nombre = stri_count(text, fixed = "$"))

  res <- locuteur
  # ↑↑↑

  df[[j]] <- res
}

df <- bind_rows(df)
df
```

Calculate the total duration for each number of speakers across the entire
collection.

```{r}
df |>
  group_by(locuteur_nombre) |>
  summarise(
    duree = sum(xmax - xmin)
  )
```

## 12. Audio Data

Load a wav format file with the code below.

```{r}
# keep
w <- readWave("donnees/phrase.wav")
snd <- makesound(w@left, fs = w@samp.rate)
snd
```

Create a `formants` dataset using the `formanttrack` function.

```{r}
formants <- formanttrack(snd, fs = fs)
```

Convert the formants data to a tibble.

```{r}
formants <- as_tibble(formants)
formants
```

I used Praat to locate the first vowel. It appears between 300 and 350
milliseconds. Visualize all the formant values of this vowel (an /a/) using code
similar to that in the earlier sections.

```{r}
formants |>
  filter(time > 300 & time < 350) |>
  ggplot() +
  geom_point(aes(f2, f1)) +
  scale_x_reverse(limits = c(500, 3500)) +
  scale_y_reverse(limits = c(250, 1400))
```

## 13. Grammatical Analysis

Load the data with the code below.

```{r}
# keep
wikifr <- read_csv2("donnees/wiki_parsed.csv.bz2")
wikifr
```

Select only the rows where the dependency relation (`dep`) equals "det"
(determiner). Select only those whose lemmas are "le" and "un". Then, grouped
by the head word (`head`), compute the mean proportion of lemmas equal to "le".
Filter to head words appearing at least 600 times and sort the results in
descending order.

```{r}
wikifr |>
  filter(dep == "det") |>
  filter(lemma %in% c("le", "un")) |>
  group_by(head) |>
  summarise(
    prop_le = mean(lemma == "le"),
    n = n()
  ) |>
  filter(n > 600) |>
  arrange(desc(prop_le))
```

## 14. PCA + UMAP

Load the data with the code below.

```{r}
# keep
pays <- read_csv2("donnees/pays.csv")
pays
```

Load the embeddings with:

```{r}
# keep
embed <- read_rds("donnees/fasttext_embed.rds")
idx <- match(pays$pays, rownames(embed))
X <- embed[idx, ]
dim(X)
```

Create the PCA components and add them to the `pays` object.

```{r}
apc <- prcomp(X, center = TRUE, scale. = TRUE)
pays$apc1 <- apc$x[, 1]
pays$apc2 <- apc$x[, 2]
pays
```

Visualize the PCA components as in the notes, with colour linked to `region`.

```{r}
pays |>
  ggplot(aes(apc1, apc2)) +
    geom_point(aes(colour = region)) +
    geom_text(
      aes(label = pays, colour = region),
      size = 2,
      nudge_y = -0.25
    )
```

Create a new table `europe` containing European countries. Then create the PCA
components, add them to the table, and visualize with the `sous_region` column.

```{r}
europe <- pays |>
  filter(region == "Europe")

embed <- read_rds("donnees/fasttext_embed.rds")
idx <- match(europe$pays, rownames(embed))
X <- embed[idx, ]
dim(X)

apc <- prcomp(X, center = TRUE, scale. = TRUE)
europe$apc1 <- apc$x[, 1]
europe$apc2 <- apc$x[, 2]

europe |>
  ggplot(aes(apc1, apc2)) +
    geom_point(aes(colour = sous_region)) +
    geom_text(
      aes(label = pays, colour = sous_region),
      size = 2,
      nudge_y = -0.25
    )
```

Create the UMAP embeddings and add them to the `europe` table. Visualize again
in the same way.

```{r}
obj <- umap(X)
europe$umap1 <- obj$layout[, 1]
europe$umap2 <- obj$layout[, 2]

europe |>
  ggplot(aes(umap1, umap2)) +
    geom_point(aes(colour = sous_region)) +
    geom_text(
      aes(label = pays, colour = sous_region),
      size = 2,
      nudge_y = -0.1
    )
```

## 15. Predictive Models

Load the data with the code below.

```{r}
# keep
parse <- read_csv2("donnees/polarity_parse.csv.bz2")
meta <- read_csv2("donnees/polarity_meta.csv.bz2")
```

Create an object `X` as in the notes but use the `pos` column instead of
`lemma`.

```{r}
X <- parse |>
  to_sparse(doc_id, pos)
dim(X)
```

Create the objects `df_meta`, `ind_train`, and `ind_valid` using the code below.

```{r}
# keep
df_meta <- tibble(doc_id = as.numeric(rownames(X))) |>
  left_join(meta, by = "doc_id")
df_meta

ind_train <- sort(sample(seq_len(nrow(X)), 10000))
ind_valid <- setdiff(seq_len(nrow(X)), ind_train)
```

Create a predictive model with the data.

```{r}
model <- cv.glmnet(
  X[ind_train,], df_meta$polarity[ind_train], family="binomial"
)
```

Evaluate the classification rate on the validation data.

```{r}
pred <- predict(model, newx=X[ind_valid,], type = "class")
mean(pred == df_meta$polarity[ind_valid])
```

Create a confusion matrix.

```{r}
table(pred=pred, y=df_meta$polarity[ind_valid])
```

Examine the model coefficients. You can pass the argument `s = model$lambda.1se`
to the `coef` function.

```{r}
cf <- coef(model, s = model$lambda.1se)
cf <- cf[cf[,1] != 0,,drop=FALSE][-1,,drop=FALSE]
cf[order(cf[,1]),,drop=FALSE]
```

## 16. LLM

Load the data with the code below and create the objects `X`, `df_meta`,
`ind_train`, and `ind_valid`.

```{r}
# keep
llm <- read_csv2("donnees/polarity_llm.csv.bz2")
llm

X <- parse |>
  mutate(lemma = stri_trans_tolower(lemma)) |>
  group_by(lemma) |>
  filter(n() > 200) |>
  to_sparse(doc_id, lemma)
df_meta <- tibble(doc_id = as.numeric(rownames(X))) |>
  left_join(meta, by = "doc_id")
ind_train <- sort(sample(seq_len(nrow(X)), 10000))
ind_valid <- setdiff(seq_len(nrow(X)), ind_train)
```

Create a predictive model for the `certain` values.

```{r}
X_nouv <- X[as.character(llm$doc_id),]
model <- cv.glmnet(X_nouv, llm$certain)
```

Examine the model coefficients.

```{r}
cf <- coef(model, s = model$lambda[12])
cf <- cf[cf[,1] != 0,,drop=FALSE][-1,,drop=FALSE]
cf[order(cf[,1]),,drop=FALSE]
```

You can repeat this with the other columns of `llm`.

## 17. Whisper

Load the data with the code below.

```{r}
# keep
whisper_mots <- read_csv2("donnees/whisper_mots.csv")
whisper_mots
```

We will study this table in section 19.

## 18. Text Embeddings

Load the data with the code below and create the objects `ind_train` and
`ind_valid`.

```{r}
# keep
X <- read_rds("donnees/polarity_embed.rds")
ind_train <- sort(sample(seq_len(nrow(X)), 700))
ind_valid <- setdiff(seq_len(nrow(X)), ind_train)
```

Create a model for `certain` as in the notes.

```{r}
model <- cv.glmnet(
  X[ind_train,], llm$certain[ind_train], alpha=0
)
```

Run the code below to create predictions.

```{r}
# keep
df <- tibble(
  certain = llm$certain[ind_valid],
  pred = as.numeric(predict(model, newx=X[ind_valid,], type = "class"))
)
df
```

Visualize the relationship between the `certain` values and the predictions.

```{r}
df |>
  ggplot() +
  geom_point(aes(factor(certain), pred))
```


## 19. Text Alignment

Load the data with the code below and create a table `tg_mots`.

```{r}
# keep
tg <- read_textgrid(
  "donnees/rhapsodie/tg/Rhap-D0001-Pro.TextGrid",
  encoding="UTF-8"
)
tg_mots <- tg |>
  filter(tier_name == "words") |>
  filter(text != "_") |>
  select(text, xmin, xmax)
```

Use the `alignement_des_textes` function to align `tg_mots` and `whisper_mots`.
Save the data in an object `df_comb`.

```{r}
df_comb <- alignement_des_textes(
  tg_mots,
  whisper_mots,
  col1 = text,
  col2 = word,
  suffix_y = "_whisper"
)
df_comb
```

Finish by calculating the mean differences for `xmin` and `xmax` between the
two datasets. You can use the `abs()` function.

```{r}
df_comb |>
  mutate(
    diff_min = xmin - start_whisper,
    diff_max = xmax - end_whisper
  ) |>
  summarise(
    diff_min_moyenne = mean(abs(diff_min)),
    diff_max_moyenne = mean(abs(diff_max))
  )
```
